Synthesizing Symmetric Lenses

### 2018-07-02 papers/SymmetricLenses/main.tex
add eval and conclusion and do a big revision

%% \bcp{This bit feels more detailed / technical than needed for an abstract.}
%%   We develop a type-directed synthesis algorithm for generating lenses in this
%%   language from regular expressions describing the data formats. A significant
%%   technical challenge in this work arises from the fact that our target language
%%   permits synchronization between formats that don't have exactly the same data
%%   -- data in one format need not be present in the other format. This gives the
%%   synthesis algorithm much more freedom, as data can be completely unaligned
%%   (edits in one format are never propagated), completely aligned (the formats
%%   are in bijective correspondence), or anywhere in between. Under the assumption
%%   that users typically want to align as much data as possible, we provide an
%%   information theoretic metric for how much data is left unaligned based on the
%%   theory of \emph{stochastic regular expressions}. Furthermore, we use this
%%   metric to develop a greedy synthesis algorithm that tries to minimize
%%   unaligned data. We experimentally demonstrate that our algorithm can
%%   efficiently synthesize simple symmetric lenses on a variety of benchmark
%%   suites consisting of synchronizing file formats from configuration files,
%%   application-specific data storage files, and data cleaning tasks.

@@

\noindent
\dpw{XXXXXX New attempt at intro starts here}

In a big-data world, one will often find similar data stored in a
variety of different formats. For instance, electronic calendars come in iCalendar, vCalendar and h-event
formats~\cite{calendar-formats}; US taxation data can be transmitted via Tax XML (used by the US government
for e-filing) or TXF (used by TurboTax)~\cite{tax-formats}; and sewing machines can read embroidery files
transmitted as PES formats, DST formats or several others~\cite{embroidery-formats}. \dpw{Editorial comment:
Phil Wadler forced his very irritating long-form citations into the
standard format. ug. how stupid is that.}
\ksf{I like the embroidery example and I also hate the long form citations.}

@@

Lens languages, like any domain-specific language, also cut down the space of
possible programs one can write, and the smaller the space of programs,
the easier it is to find the program one is looking for.  
%This feature can be a negative, if the desired transformation
%falls outside of the scope provided by the language, but can also be positive, as complex
%properties (like invertability) can be enforced automatically.
%Another advantage of reducing the
%set of programs is that doing so can facilitate program synthesis, which is essentially
%a search through the space of possible computations that can be expressed as a lens.
Recent research~\cite{optician}
has exploited this fact by demonstrating that it is possible to synthesize relatively
large, complex lenses from specifications.
In particular, given types that describe source and target formats, there are synthesis
algorithms that will find bijective lenses~\cite{optician}
for medium-sized files, such as those housing linux
application configurations.  \ksf{why medium sized?  You can have very
large files with simple format descriptions that are bijective and for
which our previous systems would have worked just fine.} Oftentimes, these algorithms find such lenses in less than
a second, and without the aid of any examples at all, with just types describing the source and
target formats.  \dpw{I would cite the ICFP paper (cite string: maina+:quotient-synthesis) as to appear but I'm not sure how to do that
without possibly violating the spirit of double-blind reviewing. ie, if it hasn't appeared yet,
how could we know the content if we aren't the authors?  Mentioning it in the 3rd person does not
seem to suffice ...} \ksf{The ICFP 2018 website lists the paper; I
think it is okay to cite it.  The reviewer can assume we contacted the
authors for a copy.  We could ask Stephanie for a ruling if we want to
be sure.  One of the principles of double blind as adopted by SIGPLAN
is not to reduce the quality of the paper; omitting this citation
would have that affect.}\bcp{+1}\saz{Stephanie says we should cite it in the third
person, even if that's a bit awkward, but otherwise citing it is OK.}

One of reasons these algorithms are so effective, even relative to successful
synthesis projects in other domains, is because unless the same kind of data
appears in a format many times, there simply are not very many different
bijections between related data formats. Hence, if the synthesis algorithm can
find any bijection at all, then it is very likely to be the one the user wants.
However, there is also a price for narrowing the synthesis domain so
dramatically: Many real data formats are almost, but not quite, bijections. More
often two related data formats have some data in common, but also contain some
differences. For instance, two related tables might share four of five columns,
but not the last. Alternately, one ad hoc configuration file might host some
format-specific meta data, such as a date, a time or reference number, whereas
the other does not. Indeed, all of our benchmarks taken from Flash
Fill~\cite{flashfill} and all of our benchmarks that synchronize file formats
contain data present in one format that isn't present in the other.
Consequently, past tools that search for strict bijections will rarely be
applicable in practice.

@@

The first difficulty was pragmatic:  symmetric lenses actually
usually operate over three structures, the source $S$, the target $T$ and a
complement $C$~\cite{symmetric-lenses}.
The latter object contains all the information not present in one of $S$ or $T$.  Logistically,
the complement object must be stored away somewhere and infrastructure must be developed
to manage it.  
Perhaps more relevant though, is that complements complicate specifications, even when
those specifications are just examples.  The issues is that one cannot
just give examples of source and target; one must also give guidance on how to manage
the ``hidden state.''  In order to avoid such complexities, we defined a new
variant of symmetric lenses, which we call \emph{simple symmetric lenses}.  Our new
lenses do not require complements; they only need source and target
data as inputs. \ksf{This seems magic.  Say something about how it is
possible/under what circumstances it is possible to not need a
complement?
At the price of some expressiveness, our new lenses eliminate the need
for complements; they require only source and target data as inputs.
}

In order to characterize the expressive power of simple symmetric lenses
theoretically, we prove that simple symmetric lenses
include all (and only) those symmetric lenses that satisfy
an intuitive property called \emph{forgetfulness}.  In other words, they are the
complete set of symmetric lenses that do not require a ``memory'' to retain data
from past instances of $S$ or $T$ when making a round trip.\dpw{fix that to make the intuitive
explaination sufficiently accurate}
From a pragmatic perspective, we justify our design choices by extending
the Boomerang implementation~\cite{Boomerang} with symmetric lenses (which
had never before been implemented\bcp{Confusing: we implement only simple
  ones, but it was the earlier full form that had never been implemented}),
applying them successfully to a range of real-world applications,
and demonstrating that they are compatible, not just with basic bijective
combinators,\bcp{confusing: they are not exactly ``compatible'' with the
  bijective combinators---they enrich them}
but with a rich set of advanced features including quotient lenses and matching lenses.

The second difficulty with synthesizing symmetric lenses is more fundamental\dpw{better word?}:
While the number of bijective lenses
between two related formats is often tiny, the number of symmetric lenses is often
large.  In practice, the difference in cardinality means that when a search algorithm selects the smallest
bijective lens it finds, that lens has a good chance of being the
one a user wants.  In contrast, and to our surprise, our experiments show \dpw{can we have an experiment that
shows this?} that
when a search algorithm selects the smallest symmetric lens, it typically fails
to find the lens the user wants.  The flexibity these lenses admit when it comes to
eliding information generates an explosion of possible transformations.
To disambiguate between the many possible lenses, we need
new principles for identifying the ``likely'' ones, and we need
new algorithms to search the space of program transformations using these principles.
\dpw{Were the bijective lenses chosen using a ``smaller'' criteria?  I
forget... if not edit the prior paragraph.}\dpw{Ideally, we could back up the claim
that there are many symmetric lenses and not many bijective lenses by graphing the
number of such lenses somehow.  eg, in a past synthesis paper with Steve (POPL 2016)
we graphed
the untyped terms of type $\mathsf{int}\rightarrow\mathsf{int}$.  I wonder if we can
do something similar here.}

To overcome this second difficulty, we turned to information theory
to try to find a way to characterize the likely and unlikely lenses.  Intuitively, a ``likely'' lens
is one that propagates a lot of information from source to target and also from target
back to source.  Conversely, given the source, the target is ``unlikely'' to require very much
additional information in order to reconstruct it---if it were\bcp{were
  what?  likely, or unlikely?}, then one is unlikely to
be attempting to synthesize a lens between the two formats\bcp{why? even if
  two big formats only share a little bit of common information, mightn't
  one want a lens between them?}!  In order to implement
this intuition, we turn to \emph{stochastic regular expressions} (SREs).
Stochastic regular expressions define a set of strings, as do ordinary regular expressions,
but they also provide the expected distribution of strings within that set.  Such
distributions allow us to compute the amount of information gained when presented
an element of the set.  For example, the language of the SRE $a |_{0.5} b$ contains two elements
(the strings $a$ and $b$).  The probability 0.5 annotating the union indicates that both
elements are equally likely to appear.  Consequently, if we observe 
an element of that language (either $a$ or $b$), we will gain one bit of information.  Said another way,
one bit, together with the SRE, suffices to recover the string in
question.\bcp{Is this implying that, if the annotation were
  different, we'd need something other than one bit to recover the string??}
On the other hand, consider the language of the SRE $c$.
It contains just one element (the string $c$).  If we observe that our algorithm
generates $c$, we gain zero bits of information --- the string was already determined.
Hence\bcp{?}, a lens is ``likely,'' in a technical sense, if relative to the target SRE $T$,
very little information from $t$ is required to generate it.

With simple symmetric lenses and SREs in hand, we develop a new algorithm for synthesizing likely lenses.
This algorithm is constructed from a pair of collaborating search procedures.
Given source and target SREs $S$ and $T$, the first algorithm searches for two \emph{related} SREs
$S'$ and $T'$ that are \emph{compatible} (a heuristic that estimates the likelihood
our second algorithm will succeed).  $S'$ and $T'$ are related in the sense
that they define the same regular sets as $S$ and $T$ (respectively) and the distribution
of strings is identical.  Indeed, we guarantee these two properties by implementing
a search procedure that obeys the star semi-ring axioms of regular expression equivalence,
and which we show does not alter the probability distributions of the languages.

Given the two compatible SREs,
the second algorithm uses a greedy, type-directed algorithm find a likely lens between them.
If algorithm estimates that the lens it has discovered is more likely than
any future lens it can possibly find, the search terminates.  Otherwise,
this lens is stored away (if it is the best found so far) and the search continues with
algorithm 1 again.

@@

In summary, our main contributions are:\bcp{annotate with section numbers}
\begin{itemize}
\item We define the space of simple symmetric lenses, which lies between asymmetric lenses and full symmetric lenses and has the useful property that they don't have any internal state.  We characterize
the simple lenses semantically as the subset of full symmetric lenses for which a ``forgetfulness''
property holds. We also analyze the gap between simple and full symmetric lenses empirically and argue that
simple ones suffice for many applications.
\item We show how to synthesize simple symmetric lenses (including all asymmetric lenses as a special case),
using a novel application of stochastic regular expressions to guide the search process.  In particular, we
show that the RE star semi-ring equivalences remain valid when their semantics are enriched by distributions.
\item We extend the Boomerang implementation with simple symmetric lenses,
  showing that they can easily be combined with a rich variety of other lens
  extensions including quotients and matching lenses. We evaluate our
  implementation over a set of benchmarks drawn from data cleaning, information
  extraction, and file synchronization tasks and demonstrate that we can
  synthesize all of them in under half a minute.
\end{itemize}

@@

\noindent
\dpw{XXXXXX New attempt at intro ends here}

% state of the world
\bcp{Not very compelling:}Similar data can come in varieties of formats: scheduled jobs can be stored as
cron jobs for Linux or as Scheduled Tasks for Windows; web service
resources can be provided to clients in legacy or modern formats;
calendar events can be stored in CSV files or ICS files. Oftentimes,
files should be synchronized across these formats: when I edit a task
to run more often on my Linux machine, I want this task to run more
often on my Windows machine; when I update data through a legacy API,
I want those changes to be reflected when I retrieve it through a
modern API; when I create a new calendar event in my ICS file, I want
my CSV calendar files to be updated to reflect this change.

Unfortunately, writing such synchronizers is quite difficult; we must write
parsers both formats, specify how the parsed data is converted into the other
format, and specify how to propagate updates from one format to the other.
Bidirectional languages ease the difficulty of writing such synchronizers,
because a single, well-typed program expresses a \emph{lens}---a group of
conversion functions guaranteed to satisfy round-tripping laws. However,
bidirectional languages are themselves difficult to program in, often requiring
users to permute data in fiddly ways while satisfying the type system's complex
unambiguity constraints.

A better approach is to \emph{synthesize} such conversion functions. However,
prior work on synthesizing lenses assumes a bijection between the source and
target, which is a severe limitation in practice. Oftentimes, fields present in
one format are not present in the other: the commands scheduled to run on Linux
machines are typically different than the commands scheduled to run on Windows
machines; updated APIs can expose new information, or omit legacy information;
ICS files contain a URL field that Google Calendar CSV files do not expose.

\emph{Symmetric lenses}~\cite{symmetric-lenses} formalize this richer class of
non-bijective conversion functions by using additional state, called the
\emph{complement}, to help with synchronization. While increasing
expressiveness, the existence of the complement makes symmetric lenses difficult
to use in real-world synchronization tasks, because lens programs must maintain
the synchronized files as well as the complements. The complement further
complicates synthesis, because input-output examples no longer suffice; instead
edit sequences must be provided.

In this paper, we contrain symmetric lenses to \emph{simple symmetric lenses}.
These constrained lenses do not use a complement, and only use the files that we
aim to synchronize as state. We identify a property, \emph{forgetfulness}, that
identifies the subset of symmetric lenses that can be written as simple
symmetric lenses. We provide combinators for building simple symmetric lenses on
string data, and integrate these combinators with Boomerang, a language for
bidirectional programming on strings. Both Boomerang and our extension type lens
terms between pairs of regular expressions. The typing judgment $\Lens : \Regex
\Leftrightarrow \RegexAlt$ means the lens term $\Lens$ encodes synchronizing
functions between the format described by $\Regex$ and the format described by
$\RegexAlt$.

We provide a tool for synthesizing simple symmetric lenses when given a suite of
input-output examples and a pair of regular expressions describing the data
formats. The synthesis algorithm uses type-directed synthesis
on the two regular expressions to synthesize a lens that is well-typed between
the provided regular expressions and satisfies the provided input-output
examples.

\saz{This paragraph is hard to follow --- maybe introduce $\ell : S
  \Leftrightarrow T$ and use the names $S$ and $T$?}\bcp{The names make it a
bit better, but it is all still rather abstract.  Indeed, the whole intro is
quite abstract---could it not be driven by an example?} A primary difficulty is
that there are a relatively many well-typed simple symmetric lenses. If we are
trying to find a lens between $\Regex$ and $\RegexAlt$, there is now a choice:
do $\Regex$ and $\RegexAlt$ represent the same data, and should be aligned
(changes to data in the language of $\Regex$ are propagated to synchronized
strings in the language of $\RegexAlt$), or should it be left unaligned (changes
to data in $\Regex$ update synchronized strings in $\RegexAlt$). Or instead
should certain components of $\Regex$ be aligned with certain components of
$\RegexAlt$, with the other components left unaligned.

Operating under the assumption that users wish to align as much data as
possible, our algorithm tries to align as much data as the types and examples
allow. This task gets harder when there are more complex decisions involved. For
example, if there are two fields in one format, and one in the other, which field
in the first format should be aligned to the sole field of the second format? Or
would it be better for the field on the right to be aligneded to a mish-mash
of subcomponents from each field?\bcp{Wouldn't these questions be answered
  by examples?} To answer these questions, we develop a cost
metric for reasoning about which lens aligns the most information.

In particular, the \emph{cost} of a lens is the expected number of bits required to
recover a string in one format when a synchronized string in the other format is
provided. In this model, bijective lenses have zero cost, and as more
information is left unsychronized, the cost is increased.

Our model is based on \emph{stochastic regular expressions}~\cite{?}, which
represent both a language and a probability distribution over that language
simultaneously.  We develop a way to syntactically infer the expected
information content (or \emph{entropy}) of an unambiguous regular
expression. This entropy is used to calculate the expected information content
to recover a string, when given a string it is synchronized with.\bcp{I
know we're going to get to more technical detail later, but at this point I
am confused: a string matching a RE can in general be arbitrarily long, so
what kind of sense does it make to talk about its ``expected'' information content?}

\begin{figure}
  \includegraphics[width=.5\textwidth]{high-level-algorithm.pdf}
  \caption{Schematic Diagram for \SOptician. Regular expressions \Regex and
    \RegexAlt, and a set of examples \Examples, are provided as input. \RXSearch
    searches through stochastic regular expression pairs, equivalent to the
    originals, and proposes them to \GreedySynth. \GreedySynth finds a lens,
    typed between the generated equivalent pairs. When \RXSearch determines it
    has an optimal lens, it returns that lens.}
  \label{fig:high-level-algorithm}
\end{figure}

\bcp{It gets rather technical and incomprehensible for the next three
  paragraphs...}

We develop an algorithm that synthesizes a low-cost lens, shown in
Figure~\ref{fig:high-level-algorithm}. This algorithm consists of two
communicating synthesizers, one that proposes candidate stochastic regular
expression pairs that it thinks will align well, and one that greedily searches
for a lens between those two regular expressions that minimizes cost and does
not traverse star-related equivalences\bcp{???}. When the first algorithm is content with
the lens it has found, it returns that lens.

\bcp{Make clear that these algorithms are based on earlier / published ones.}Our first algorithm, \RXSearch, proposes candidate stochastic regular
expressions. These candidate stochastic regular expressions are generated by
traversing \emph{star semiring} equivalences on stochastic regular expressions.
These proposed stochastic regular expression pairs have equivalent languages and
probability distributions as the original ones: we prove that the star-semiring
rewrites this synthesizer applies are sound rewrites on stochastic regular
expressions.

Our second algorithm, \GreedySynth, utilizes alternative languages for
stochastic regular expressions and lenses, \emph{stochastic DNF regular
  expressions} and \emph{symmetric DNF lenses}. While these languages are hard
to read and understand, they are highly normalized and permit an efficient
type-directed synthesis algorithm. We develop our cost metric on DNF lenses, and
provide a greedy algorithm that tries to find a symmetric DNF lens of minimal
cost. The synthesized symmetric DNF lens is then converted to a lens in the
original simple symmetric lens language.

We evaluate our algorithm on a benchmark suite consisting of synchronizing file
formats from configuration files, application-specific data storage files, and
data cleaning tasks. We find that our algorithm is able to synthesize each of
these lenses in under 20 seconds.

In summary, our contributions are:
\begin{enumerate}
\item We develop \emph{simple symmetric lenses} a new formulation for
  bidirectional programs that can project data from both sides~(\S\ref{sec:overview}),
  and describe a language that expresses such lenses on string
  data~(\S\ref{sec:ssl}).  \bcp{A new language, or a variant of
    Hofmann/Pierce/Wagner?} 
\item We develop a means to traverse stochastic regular expression equivalences,
  while preserving probability distributions and
  languages~(\S\ref{subsec:stoch-rx}).
\item We develop stochastic DNF regular expressions and DNF
  lenses~(\S\ref{subsec:greedy-synth}). We develop a sound algorithm for
  converting stochastic regular expressions to stochastic DNF regular
  expressions and a information-theoretic cost metric on DNF lenses.
\item We extended Boomerang with our simple symmetric lens combinators, and
  integrated our algorithm with the Optician synthesis system within Boomerang,
  allowing synthesis tasks as Boomerang
  expressions~(\S\ref{sec:implementation}).
\item We evaluate our algorithm on a benchmark suite consisting of synchronizing
  file formats from configuration files, application-specific data storage
  files, and data cleaning tasks (\S\ref{sec:evaluation}).  \bcp{... and
    conclude what?}
\end{enumerate}
% end introduction

@@

\ksf{Commented out entropy calculation; I don't think we need that
level of detail; leaving the text in case we change our mind :-)}
%Not sure we need this level of detail after all.
%\afm{calculations for cost of various lenses between \lstinline{emp_salary} and
%  \lstinline{emp_row}} Consider the lens \lstinline{single_emp_map} between
%\lstinline{emp_salary} and \lstinline{emp_row}. Given this lens, the expected
%entropy of recovering the left format from the right
%($\EntropyOf{\lstinline{emp_salary} \Given \lstinline{emp_row},
%  \lstinline{single_emp_map} }$) is the sum of the expected entropy of
%recovering each of its subcomponents, $\EntropyOf{\lstinline{emp_salary} \Given
%  \lstinline{emp_row}, \lstinline{single_emp_map} } = \EntropyOf{\lstinline{name
%    . " " . name} \Given \lstinline{name . "," . name}, \lstinline{name_map}
%}$\\ + $\EntropyOf{\lstinline{": " . salary} \Given
%  \lstinline{"," . company}, \lstinline{disc(salary,company,"UNSET","UNSET")}
%}$.
%
%The entropy of the identity lens is zero. As the two formats are in bijective
%correspondence, we know exactly which string is present in the left-hand format.
%The entropy of the disconnect lens is just the entropy of
%\lstinline{": " . salary}.  Knowing which string is present in the right-hand
%format tells us nothing about what string is present in the left-hand format. So
%$\EntropyOf{\lstinline{emp_salary} \Given \lstinline{emp_row},
%  \lstinline{single_emp_map} } = \EntropyOf{\lstinline{": " . salary} }$
%
%Analogously, $\EntropyOf{\lstinline{emp_row} \Given \lstinline{emp_salary},
%  \lstinline{single_emp_map} }$ is just $\EntropyOf{\lstinline{"," . company}
%}$, so the total cost of the lens is $\EntropyOf{\lstinline{": " . salary}} +
%\EntropyOf{\lstinline{"," . company}}$

@@

% begin ssl
\section{Simple Symmetric String Lenses}
\label{sec:ssl}
\sam{Perhaps we shoud introduce stochastic regular expressions first before introducing the lens combinators? This is because the type equivalence combinator mentions regexp equivalence, but this has not been introduced by this point in the paper.}
We now give a formal description of all the combinators in our symmetric lens
language. These lens terms are typed between pairs of regular expressions: the
typing judgment $\Lens \OfType \Regex \Leftrightarrow \RegexAlt$ means that
$\Lens$ is a well typed lens between $\LanguageOf{\Regex}$ and
$\LanguageOf{\RegexAlt}$.

>>>>

%% \bcp{This bit feels more detailed / technical than needed for an abstract.}
%%   We develop a type-directed synthesis algorithm for generating lenses in this
%%   language from regular expressions describing the data formats. A significant
%%   technical challenge in this work arises from the fact that our target language
%%   permits synchronization between formats that don't have exactly the same data
%%   -- data in one format need not be present in the other format. This gives the
%%   synthesis algorithm much more freedom, as data can be completely unaligned
%%   (edits in one format are never propagated), completely aligned (the formats
%%   are in bijective correspondence), or anywhere in between. Under the assumption
%%   that users typically want to align as much data as possible, we provide an
%%   information theoretic metric for how much data is left unaligned based on the
%%   theory of \emph{stochastic regular expressions}. Furthermore, we use this
%%   metric to develop a greedy synthesis algorithm that tries to minimize
%%   unaligned data. We experimentally demonstrate that our algorithm can
%%   efficiently synthesize simple symmetric lenses on a variety of benchmark
%%   suites consisting of synchronizing file formats from configuration files,
%%   application-specific data storage files, and data cleaning tasks.
TODO: abstract

@@

\noindent
In a big-data world, one will often find similar data stored in a
variety of different formats. For instance, electronic calendars come in iCalendar, vCalendar and h-event
formats~\cite{calendar-formats}; US taxation data can be transmitted via Tax XML (used by the US government
for e-filing) or TXF (used by TurboTax)~\cite{tax-formats}; and sewing machines can read embroidery files
transmitted as PES formats, DST formats or several others~\cite{embroidery-formats}. \dpw{Editorial comment:
Phil Wadler forced his very irritating long-form citations into the
standard format. ug. how stupid is that.}
\ksf{I like the embroidery example and I also hate the long form citations.}

@@

Lens languages, like any domain-specific language, also cut down the space of
possible programs one can write, and the smaller the space of programs,
the easier it is to find the program one is looking for.  
%This feature can be a negative, if the desired transformation
%falls outside of the scope provided by the language, but can also be positive, as complex
%properties (like invertability) can be enforced automatically.
%Another advantage of reducing the
%set of programs is that doing so can facilitate program synthesis, which is essentially
%a search through the space of possible computations that can be expressed as a lens.
Recent research~\cite{optician} has exploited this fact by demonstrating that it
is possible to synthesize relatively large, complex lenses from specifications.
In particular, given types that describe source and target formats, there are
synthesis algorithms that will find bijective lenses~\cite{optician} for files
in ad-hoc formats, such as those housing linux application configurations.
Oftentimes, these algorithms find such lenses in less than a second, and without
the aid of any examples at all, with just types describing the source and target
formats. \dpw{I would cite the ICFP paper (cite string:
  maina+:quotient-synthesis) as to appear but I'm not sure how to do that
  without possibly violating the spirit of double-blind reviewing. ie, if it
  hasn't appeared yet, how could we know the content if we aren't the authors?
  Mentioning it in the 3rd person does not seem to suffice ...} \ksf{The ICFP
  2018 website lists the paper; I think it is okay to cite it. The reviewer can
  assume we contacted the authors for a copy. We could ask Stephanie for a
  ruling if we want to be sure. One of the principles of double blind as adopted
  by SIGPLAN is not to reduce the quality of the paper; omitting this citation
  would have that affect.}\bcp{+1}\saz{Stephanie says we should cite it in the
  third person, even if that's a bit awkward, but otherwise citing it is OK.}

One of reasons these algorithms are so effective, even relative to successful
synthesis projects in other domains, is because unless the same kind of data
appears in a format many times, there simply are not very many different
bijections between related data formats. Hence, if the synthesis algorithm can
find any bijection at all, then it is very likely to be the one the user wants.
However, there is also a price for narrowing the synthesis domain so
dramatically: Many real data formats are almost, but not quite, bijections. More
often two related data formats have some data in common, but also contain some
differences. For instance, two related tables might share four of five columns,
but not the last. Alternately, one ad hoc configuration file might host some
format-specific meta data, such as a date, a time or reference number, whereas
the other does not. Indeed, all of our benchmarks taken from Flash
Fill~\cite{flashfill} and all of our benchmarks that synchronize between two
ad-hoc file formats. Consequently, past tools that search for strict bijections
will rarely be applicable in practice.

@@

The first difficulty was pragmatic: symmetric lenses actually usually operate
over three structures, the source $S$, the target $T$ and a complement
$C$~\cite{symmetric-lenses}. The latter object contains all the information not
present in one of $S$ or $T$. Logistically, the complement object must be stored
away somewhere and infrastructure must be developed to manage it. Perhaps more
relevant though, is that complements complicate specifications, even when those
specifications are just examples. The issues is that one cannot just give
examples of source and target; one must also give guidance on how to manage the
``hidden state.'' In order to avoid such complexities, we defined a new variant
of symmetric lenses, which we call \emph{simple symmetric lenses}. At the price
of some expressiveness, our new lenses do not require complements; they only
need source and target data as inputs.

In order to characterize the expressive power of simple symmetric lenses
theoretically, we prove that simple symmetric lenses include all (and only)
those symmetric lenses that satisfy an intuitive property called
\emph{forgetfulness}. In other words, they are the complete set of symmetric
lenses that do not require external ``memory'' to recover data from past
instances of $S$ or $T$ when making a round trip -- they only need the most
recent instance. From a pragmatic perspective, we justify our design choices by
extending the Boomerang implementation~\cite{Boomerang} with simple symmetric
lenses applying them successfully to a range of real-world applications, and
demonstrating that they can enrich, not just with basic bijective combinators,
but with a rich set of advanced features including quotient lenses and matching
lenses.

The second difficulty with synthesizing symmetric lenses is more
fundamental\dpw{better word?}: While the number of bijective lenses between two
related formats is often tiny, the number of symmetric lenses is often large. In
practice, the difference in cardinality means that when a search algorithm
selects the first bijective lens it finds, that lens has a good chance of being
the one a user wants. In contrast, and to our surprise, our experiments show
that if our search algorithm selects the first symmetric lens it finds, it
typically fails to find the lens the user wants. The flexibity these lenses
admit when it comes to eliding information generates an explosion of possible
transformations. To disambiguate between the many possible lenses, we need new
principles for identifying the ``likely'' ones, and we need new algorithms to
search the space of program transformations using these principles. \dpw{Were
  the bijective lenses chosen using a ``smaller'' criteria? I forget... if not
  edit the prior paragraph.}\afm{no, chosen using a ``first'' criteria.
  Edited.}\dpw{Ideally, we could back up the claim that there are many symmetric
  lenses and not many bijective lenses by graphing the number of such lenses
  somehow. eg, in a past synthesis paper with Steve (POPL 2016) we graphed the
  untyped terms of type $\mathsf{int}\rightarrow\mathsf{int}$. I wonder if we
  can do something similar here.}\afm{What about lenses between ``a''* and
  ''b``*}

To overcome this second difficulty, we turned to information theory to try to
find a way to characterize the likely and unlikely lenses. Intuitively, a
``likely'' lens is one that propagates a lot of information from source to
target and also from target back to source. Conversely, given the source, the
target is ``unlikely'' to require very much additional information in order to
reconstruct it---if it were\bcp{were what? likely, or unlikely?}, then one is
unlikely to be attempting to synthesize a lens between the two formats\bcp{why?
  even if two big formats only share a little bit of common information,
  mightn't one want a lens between them?}! In order to implement this intuition,
we turn to \emph{stochastic regular expressions} (SREs). Stochastic regular
expressions define a set of strings, as do ordinary regular expressions, but
they also provide the expected distribution of strings within that set. Such
distributions allow us to compute the amount of information gained when
presented an element of the set. For example, the language of the SRE
$\lstinline{"a"} |_{0.5}
\lstinline{"b"}$ contains two elements (the strings
\lstinline{"a"} and
\lstinline{"b"}). The probability 0.5 annotating the union indicates that both
elements are equally likely to appear. Consequently, if we observe an element of
that language (either \lstinline{"a"} or \lstinline{"b"}), we will gain one bit
of information. Said another way, one bit, together with the SRE, suffices to
recover the string in question.\bcp{Is this implying that, if the annotation
  were different, we'd need something other than one bit to recover the
  string??}
\afm{From an information theoretic perspective, yes, because that perspective
  views it as the amount of information if we were communicating a stream of
  these choices.} On the other hand, consider the language of the SRE \lstinline{c}. It contains
just one element (the string $c$). If we observe that our algorithm generates
$\lstinline{"c"}$, we gain zero bits of information --- the string was already determined.
Hence\bcp{?}, a lens is ``likely,'' in a technical sense, if relative to the
target SRE $T$, very little information from $t$ is required to generate it.

\begin{figure}
  \includegraphics[width=.5\textwidth]{high-level-algorithm.pdf}
  \caption{Schematic Diagram for our symmetric lens synthesis algorithm. Regular
    expressions \Regex and \RegexAlt, and a set of examples \Examples, are
    provided as input. \RXSearch searches through stochastic regular expression
    pairs, equivalent to the originals, and proposes them to \GreedySynth.
    \GreedySynth finds a lens, typed between the generated equivalent pairs.
    When \RXSearch determines it has an optimal lens, it returns that lens.}
  \label{fig:high-level-algorithm}
\end{figure}

With simple symmetric lenses and SREs in hand, we develop a new algorithm for
synthesizing likely lenses, based on the previous algorithm for synthesizing
bijective lenses. This algorithm is constructed from \RXSearch and \GreedySynth,
a pair of collaborating search procedures, shown in
Figure~\ref{fig:high-level-algorithm}. Given source and target SREs $S$ and $T$,
the \RXSearch searches for two \emph{related} SREs $S'$ and $T'$ that are
\emph{compatible} (a heuristic that estimates the likelihood our second
algorithm will succeed). $S'$ and $T'$ are related in the sense that they define
the same regular sets as $S$ and $T$ (respectively) and the distribution of
strings is identical. Indeed, we guarantee these two properties by implementing
a search procedure that obeys the star semi-ring axioms of regular expression
equivalence, and which we show does not alter the probability distributions of
the languages.

Given the two compatible SREs, \GreedySynth uses a greedy, type-directed
algorithm find a likely lens between them. If algorithm estimates that the lens
it has discovered is more likely than any future lens it can possibly find, the
search terminates. Otherwise, this lens is stored away (if it is the best found
so far) and the search continues with \RXSearch again.

@@

In summary, our main contributions are:\bcp{annotate with section numbers}
\begin{itemize}
\item We define the space of simple symmetric lenses, which lies between
  asymmetric lenses and full symmetric lenses and has the useful property that
  they don't have any internal state (\S\ref{sec:overview}). We characterize the
  simple lenses semantically as the subset of full symmetric lenses for which a
  ``forgetfulness'' property holds (\S\ref{sec:related}). We also analyze the
  gap between simple and full symmetric lenses empirically and argue that simple
  ones suffice for many applications (\S\ref{sec:evaluation}).
\item We show how to synthesize simple symmetric lenses (including all asymmetric lenses as a special case),
using a novel application of stochastic regular expressions to guide the search process.  In particular, we
show that the RE star semi-ring equivalences remain valid when their semantics
are enriched by distributions (\S\ref{sec:synthesis}).
\item We extend the Boomerang implementation with simple symmetric lenses,
  showing that they can easily be combined with a rich variety of other lens
  extensions including quotients and matching lenses. We evaluate our
  implementation over a set of benchmarks drawn from data cleaning, information
  extraction, and file synchronization tasks and demonstrate that we can
  synthesize all of them in under half a minute (\S\ref{sec:evaluation}).
\end{itemize}

@@

\noindent

% state of the world
% end introduction

@@

%Not sure we need this level of detail after all.
%\afm{calculations for cost of various lenses between \lstinline{emp_salary} and
%  \lstinline{emp_row}} Consider the lens \lstinline{single_emp_map} between
%\lstinline{emp_salary} and \lstinline{emp_row}. Given this lens, the expected
%entropy of recovering the left format from the right
%($\EntropyOf{\lstinline{emp_salary} \Given \lstinline{emp_row},
%  \lstinline{single_emp_map} }$) is the sum of the expected entropy of
%recovering each of its subcomponents, $\EntropyOf{\lstinline{emp_salary} \Given
%  \lstinline{emp_row}, \lstinline{single_emp_map} } = \EntropyOf{\lstinline{name
%    . " " . name} \Given \lstinline{name . "," . name}, \lstinline{name_map}
%}$\\ + $\EntropyOf{\lstinline{": " . salary} \Given
%  \lstinline{"," . company}, \lstinline{disc(salary,company,"UNSET","UNSET")}
%}$.
%
%The entropy of the identity lens is zero. As the two formats are in bijective
%correspondence, we know exactly which string is present in the left-hand format.
%The entropy of the disconnect lens is just the entropy of
%\lstinline{": " . salary}.  Knowing which string is present in the right-hand
%format tells us nothing about what string is present in the left-hand format. So
%$\EntropyOf{\lstinline{emp_salary} \Given \lstinline{emp_row},
%  \lstinline{single_emp_map} } = \EntropyOf{\lstinline{": " . salary} }$
%
%Analogously, $\EntropyOf{\lstinline{emp_row} \Given \lstinline{emp_salary},
%  \lstinline{single_emp_map} }$ is just $\EntropyOf{\lstinline{"," . company}
%}$, so the total cost of the lens is $\EntropyOf{\lstinline{": " . salary}} +
%\EntropyOf{\lstinline{"," . company}}$

@@

\section{Stochastic Regular Expressions}
\label{sec:srx}
Important to our algorithm is finding the expected number of bits needed to
recover a string in one data source from a synchronized string in the other data
source. To do this, we must first develop a probability model for our language.
We do this with \emph{stochastic regular expressions}, regular expressions
annotated with probability information. With this, we can jointly express a
language and a probability distribution over that language.

\subsection{Stochastic Regular Expression Equivalences}
$\RXSearch$ needs to be able to find equivalent stochastic regular expressions.
However, as of yet, there is no syntactic means to reason about stochastic
regular expression equivalence. We extend the \emph{star-semiring} equivalences
to stochastic regular expressions.

\subsection{Stochastic Regular Expression Entropy}
The entropy of a stochastic regular expression can be computed directly from its
syntax, when the stochastic regular expression is unambiguous and contains only
nonempty subcomponents.
\begin{center}
  \begin{tabular}{rcl}
    $\EntropyOf{s}$
    & =
    & 1\\

    $\EntropyOf{\PRegexStar{\Regex}{\Probability}}$
    & =
    & $\frac{\Probability}{1-\Probability}(\EntropyOf{\Regex} - \Log_2\Probability)
      - \Log_2(1-\Probability)$\\
